Smart Drill-Down

### 2014-11-29 TableSummarization/Paper/intro.tex
Added proof sketch of NP-hardness

To reiterate, our chief contribution in this paper is the {\em smart drill-down} interaction operator,
an extension of traditional drill-down, aimed at allowing analysts to zoom into the
more ``interesting'' parts of a dataset. In addition to this operator,
we develop techniques to support this operator on increasingly larger datasets:
\squishlist
\item {\em Basic Interaction:} We demonstrate that finding the optimal list of rules is {\sc NP-Hard},
and we develop an algorithm to find the approximately optimal list of rules to display when the user performs a smart drill-down operation. b
\item {\em Dynamic Sample Maintenance:} To improve response time on large tables, we formalize the problem of dynamically
maintaining samples in memory to support smart drill-down. 
We show that optimal identification of samples is once again {\sc NP-Hard}, \agp{??} and 
we develop an approximate scheme for dynamically maintaining and using multiple samples of the table in memory.
% to improve response time on large tables.
\squishend

>>>>

To reiterate, our chief contribution in this paper is the {\em smart drill-down} interaction operator,
an extension of traditional drill-down, aimed at allowing analysts to zoom into the
more ``interesting'' parts of a dataset. In addition to this operator,
we develop techniques to support this operator on increasingly larger datasets:
\squishlist
\item {\em Basic Interaction:} We demonstrate that finding the optimal list of rules is {\sc NP-Hard},
and we develop an algorithm to find the approximately optimal list of rules to display when the user performs a smart drill-down operation. b
\item {\em Dynamic Sample Maintenance:} To improve response time on large tables, we formalize the problem of dynamically
maintaining samples in memory to support smart drill-down. 
We show that optimal identification of samples is once again {\sc NP-Hard}, and 
we develop an approximate scheme for dynamically maintaining and using multiple samples of the table in memory.
% to improve response time on large tables.
\squishend

### 2014-11-29 TableSummarization/Paper/sampling.tex
Added proof sketch of NP-hardness

\item \textbf{Delete/Shrink:} When the SampleHandler uses \textbf{Create}, it may not have enough memory left to create a new sample. In that case, the SampleHandler needs to delete or shrink an existing sample. It could use a heuristic, like a LRU (Least-Recently-Used) policy to decide which sample to delete or shrink. Note that shrinking a sample by randomly deleting some tuples (even if the sample size goes below $minSS$) may be better than deleting it entirely, because in future, we may be able to recreate a sample using \textbf{Combine} on the existing shrinked sample along with other samples.
%For shrink, prefer shrinking samples of the largest size (> minSS) first?
\end{enumerate}

@@

At any stage, we have a tree $U$ displayed to the user, with each node of the tree corresponding to a displayed rule. The tree is formed as follows: The root of the tree corresponds to the trivial rule. And suppose the user expands a node with rule $r$, resulting in rules $r_1, r_2, .. r_k$ being displayed. Then we add children nodes to the expanded node, corresponding to rules $r_1, r_2, .. r_k$. Note that multiple nodes of the tree may correspond to the same rule. Internal nodes of $U$ are ones that have been expanded (drilled down on), while leaves are nodes that have not been expanded. Let $L$ be the set of leaves. Each leaf is something that the user can potentially expand in the next step, and thus we would like to have pre-fetched samples for rules corresponding to leaf nodes.
In the rest of this section, we abuse notation to use the term rule to refer to any node that corresponds to that rule.

We assume that we have a probability distribution over leaves, which assigns a probability that each leaf will be the next one to be expanded. In the absence of additional information, we can assume a uniform probability distribution. That is, we can assume that every leaf is equally likely to be expanded next. We can improve on this by assigning zero probability to leaves whose rule is also displayed in an internal node of the tree. We can also use Machine Learning on past user behaviour, along with node features such as `node depth in tree' and `node distance from last expanded node' to get a better probability estimate of each node being expanded next.

@@

Let the `selectivity' of a rule be the fraction of tuples in $T$ that are covered by the rule. For each pair of rules $r_1, r_2 \in U$ such that $r_1$ is a sub-rule of $r_2$, we can estimate the ratio of selectivities of $r_1$ and $r_2$ using existing samples. Call this quantity $S(r_1, r_2)$. We define $S(r_1, r_2)$ to be $0$ if $r_1$ is not a sub-rule of $r_2$. If the same rule occurs in multiple nodes $r_1$, $r_2$ of the tree, then $S(r_1, r_2)$ is naturally $1$. Then if we have an $n_r$ sized sample with filter $r$ for each $r \in U$, the expected number of tuples covered by $r^{\prime} \in L$ is given by $$\sum_{r \in U} S(r, r^{\prime})n_r$$ For any $r^{\prime}$, define the function {\em ess} (for `effective sample size') as below: 
\begin{definition}\label{def:ess}
$$ess(r^{\prime}) = \sum_{r \in U} S(r, r^{\prime})n_r$$ 
\end{definition}
If $ess(r^{\prime}) \geq minSS$ then if the user expands $r^{\prime}$, we display the next rule list using \textbf{Find} or \textbf{Combine}, instead of having to make another pass through the table. We wish to maximize the probability that we can respond to the next user expansion without making another pass. We now formally define our problem below:
\begin{problem}\label{prob:sample-sizes}
Given a tree of rules $U$ with leaves $L$, a probability distribution $p$ over $L$, an integer `capacity', and selectivity ratio $S(r_1, r_2)$ for each $r_1, r_2 \in U$, 
choose an integer $n_r \geq 0$ for each $r \in U$ so as to $\textrm{maximize}$ :
$$\sum_{r^{\prime} \in L} p_{r^{\prime}}I_{ess(r^{\prime}) \geq minSS}$$
where the $I$'s are indicator variables, subject to :
$$\sum_{r \in U} n_r < \text{capacity}$$
\end{problem}
Problem~\ref{prob:sample-sizes} is non-linear and non-convex because of the indicator variables. If the tree $U$ is not too big, then it might be feasible to use an exponential algorithm to solve the problem.

-----Hinge-loss begin-----

However, we can make the problem convex (and hence tractable) with two simplifications. The first simplification is, we modify our objective function to use hinge-loss instead of a step function. That is, our new objective function to maximise is $$\sum_{r^{\prime} \in L} p_{r^{\prime}}\textrm{min}(1, \frac{ess(r^{\prime})}{minSS} $$
Here we assume that it is acceptable to run our algorithm on samples smaller than $minSS$,  though we still prefer bigger sample sizes upto $minSS$. The other simplification we make is assuming that the sample sizes $n_r$ are real numbers instead of integers. After obtaining our optimal sample sizes, we can round them up to get integer sample sizes. This will increase the memory usage by at most $|U|$, the number of nodes in displayed tree. $|U|$ is usually negligible compared to the memory capacity, or $minSS$.

----Hinge-loss-end-----

Otherwise, with an additional simplification, we can reduce it to the knapsack-like problem, and use a PTAS (Polynomial Time Approximation Scheme) to find the approximately optimal solution. The simplification is: For each $r \in L$, we assume that it will get tuples only from samples obtained for itself (filter $= r$) and its immediate parent. That is, we set $S(r_1, r_2)$ to be zero if $r_1 \neq r_2$ and $r_2$ is not a child of $r_1$. $ess(r)$ is redefined according to the new value of $S$ as well. So $$ess(r^{\prime}) = n_{r^{\prime}} + n_rS(r, r^{\prime})$$ where $r$ is the parent of $r^{\prime}$.

@@

\techreporttext{
\subsubsection{Additional optimizations}
There are some additional minor optimizations we can make to reduce the memory cost per sample, allowing us to store more and bigger samples. 
Suppose we have a sample $s$, and say its filter rule $f_s$ has value $v$ in column $c$. Then we know that each tuple $t$ in $T_s$ must also have value $v$ in column $c$, since it is covered by $f_s$. So we do not need to explicitly store the column $c$ value of any tuple in $T_s$. We only need to store the tuple values of columns that have a $\star$ value in $f_s$.
In addition, we may have a tuple occur in multiple samples. Instead of storing the entire tuple repeatedly, we could create a dictionary of common tuples, and only store a pointer to the tuple's dictionary entry in $T_s$.

>>>>

\item \textbf{Delete/Shrink:} When the SampleHandler uses \textbf{Create}, it may not have enough memory left to create a new sample. In that case, the SampleHandler needs to delete or shrink an existing sample. It could use a heuristic, like a LRU (Least-Recently-Used) policy to decide which sample to delete or shrink. Note that shrinking a sample by randomly deleting some tuples (even if the sample size goes below $minSS$) may be better than deleting it entirely, because in future, we may be able to recreate a sample using \textbf{Combine} on the existing shrinked sample along with other samples.
\end{enumerate}

@@

At any stage, we have a tree $U$ displayed to the user, with each node of the tree corresponding to a displayed rule. The tree is formed as follows: The root of the tree corresponds to the trivial rule. And suppose the user expands a node with rule $r$, resulting in rules $r_1, r_2, .. r_k$ being displayed. Then we add children nodes to the expanded node, corresponding to rules $r_1, r_2, .. r_k$. Note that multiple nodes of the tree may correspond to the same rule. Thus even though the structure displayed to the user is a tree, the set of rules displayed forms a partially ordered set (poset) using the sub-rule ordering. Internal nodes of $U$ are ones that have been expanded (drilled down on), while leaves are nodes that have not been expanded. Let $L$ be the set of leaves. Each leaf is something that the user can potentially expand in the next step, and thus we would like to have pre-fetched samples for rules corresponding to leaf nodes.
In the rest of this section, we abuse notation to use the term rule to refer to any node that corresponds to that rule.

We assume that we have a probability distribution over leaves, which assigns a probability that each leaf will be the next one to be expanded. In the absence of additional data, we can assume a uniform probability distribution. That is, we can assume that every leaf is equally likely to be expanded next. If we have data on past user behaviour available, we can use Machine Learning on node features such as `node depth in tree', `weight of node rule' and `node distance from last expanded node' to get a better probability estimate of each node being expanded next.

@@

Let the `selectivity' of a rule be the fraction of tuples in $T$ that are covered by the rule. For each pair of rules $r_1, r_2 \in U$ such that $r_1$ is a sub-rule of $r_2$, we can estimate the ratio of selectivities of $r_1$ and $r_2$ using existing samples. Call this quantity $S(r_1, r_2)$. We define $S(r_1, r_2)$ to be $0$ if $r_1$ is not a sub-rule of $r_2$. If the same rule occurs in multiple nodes $r_1$, $r_2$ of the tree, then $S(r_1, r_2)$ is naturally $1$. Then if we have an $n_r$ sized sample with filter $r$ for each $r \in U$, the expected number of tuples covered by $r^{\prime} \in L$ is given by $$\sum_{r \in U} S(r, r^{\prime})n_r$$ For any $r^{\prime}$, define the function {\em ess} (for `effective sample size') as below: 
\begin{definition}\label{def:ess}
$$ess(r^{\prime}) = \sum_{r \in U} S(r, r^{\prime})n_r$$ 
\end{definition}
If $ess(r^{\prime}) \geq minSS$ then if the user expands $r^{\prime}$, we display the next rule list using \textbf{Find} or \textbf{Combine}, instead of having to make another pass through the table. We wish to maximize the probability that we can respond to the next user expansion without making another pass. We now formally define our problem below:
\begin{problem}\label{prob:sample-sizes}
Given a tree of rules $U$ with leaves $L$, a probability distribution $p$ over $L$, an integer `capacity', and selectivity ratio $S(r_1, r_2)$ for each $r_1, r_2 \in U$, 
choose an integer $n_r \geq 0$ for each $r \in U$ so as to $\textrm{maximize}$ :
$$\sum_{r^{\prime} \in L} p_{r^{\prime}}I_{ess(r^{\prime}) \geq minSS}$$
where the $I$'s are indicator variables, subject to :
$$\sum_{r \in U} n_r < \text{capacity}$$
\end{problem}
Problem~\ref{prob:sample-sizes} is non-linear and non-convex because of the indicator variables. If the tree $U$ is not too big, then it might be feasible to use an exponential algorithm to solve the problem. We can show that Problem~\ref{prob:sample-sizes} is NP-Hard by reducing the knapsack problem to it. 
\begin{lemma}
Problem~\ref{prob:sample-sizes} is NP-Hard.
\end{lemma}
\papertext{
\begin{proof}(Sketch)
Suppose we are given an instance of the knapsack problem with $m$ objects, with the $i^{th}$ object having weight $w_i$ and value $v_i$. We are also given a weight limit $W$, and our objective is to choose a set of objects that maximizes value and has total weight $< W$. We will reduce this to an instance of Problem~\ref{prob:sample-sizes}.

We first scale the $w_i$s and $W$ such that all $w_i$s are $< 1$, without effectively changing the problem. For Problem~\ref{prob:sample-sizes}, we set $d$ (branching factor) equal to $2$. $capacity$ is set to $(m + W) \times minSS$. Tree $U$ has $m$ special nodes $r_1, r_2, ... r_m$, and each $r_i$ has two children $r_{i,1}, r_{i,2}$. These $2m$ children are all leaves, and all leaves other than these have expansion probability $0$. 
The $S$ values are such that $\forall i \in \left\lbrace 1, 2, ..m\right\rbrace, j \in \left\lbrace 1, 2 \right\rbrace : (S(x, r_{i,j}) \neq 0 \Rightarrow x = r_{i,j} || x = r_i)$. In reality, the $S$ values cannot be exactly zero when the first argument is the trivial rule, but we can make it small enough such that any optimal solution will set $n_r = 0$ when $r$ is not a special node or its child. Thus each each leaf $r_{i,j}$ gets tuples either from its own sample, or the sample of its parent, and from nowhere else. In addition, $S(r_i, r_{i,1}) = 1$ (again, it cannot be exactly $1$, but can be brought arbitrarily close to $1$), and $S(r_i, r_{i,2}) = 1 - w_i$. Finally, for each i, $p_{r_{i,1}} = \frac{2}{2m+1}$ and $p_{r_{i,2}} = \frac{v_i}{(2m+1)\sum_{j=1}^{m}v_i}$. Thus each individual $p_{r_{i,1}}$ value is higher than all $p_{r_{i,2}}$ values combined, and $M$ is high enough to cover all $r_{i,1}$. Thus in any optimal solution, $ess(r_{i,1})$ must be $= minSS$, and we're left to decide which $i$'s should also have $ess(r_{i,2}) = minSS$. For all $i$, we must have either $n_{r_i} = n_{r_{i,2}} = 0 \land n_{r_{i,1}} = minSS$ (iff $ess(r_{i,2}) < minSS$) or $n_{r_i} = minSS \land n_{r_{i,1}} = 0 \land n_{r_{i,2}} = minSS (1 - S(r_i, r_{i,2})) = minSS \times w_i$ (iff $ess(r_{i,2}) = minSS$). The latter option consumes $w_i \times minSS$ and gives and extra $\frac{v_i}{(2m+1)(\sum_{j=1}^{m}v_j)}$ probability value. Thus having $ess(r_{i,2}) = minSS$ is equivalent to picking object $i$ from the knapsack problem (as it consumes additional memory $\propto w_i$ and gives additional probability value $\propto v_i$). Moreover the additional memory available (on top of the $m \times minSS$ required to cover all $r_{i,1}$s) is $W\times minSS$. Hence solving Problem~\ref{prob:sample-sizes} with the above $U$, $S$, $p$ and picking the set of $i$s for which $ess(r_{i,2}) = minSS$ gives a solution to the original knapsack problem. 
\end{proof}}
\techreporttext{
\begin{proof}

\end{proof}
}

However, with an additional simplifying assumption, we can make the problem approximately solvable. it to the knapsack-like problem, and use a PTAS (Polynomial Time Approximation Scheme) to find the approximately optimal solution. The simplification is: For each $r \in L$, we assume that it will get tuples only from samples obtained for itself (filter $= r$) and its immediate parent. That is, we set $S(r_1, r_2)$ to be zero if $r_1 \neq r_2$ and $r_2$ is not a child of $r_1$. $ess(r)$ is redefined according to the new value of $S$ as well. So $$ess(r^{\prime}) = n_{r^{\prime}} + n_rS(r, r^{\prime})$$ where $r$ is the parent of $r^{\prime}$.

@@

\techreporttext{
\subsubsection{Alternative Convex-Optimization based solution}
We noted earlier that Problem~\ref{prob:sample-sizes} is NP-Hard, but can be approximately solved with an additional simplifying assumption regarding the $S(r_1, r_2)$ values. Instead of making this simplification, we can make the problem convex (and hence tractable) with two different simplifications. The first simplification is, we modify our objective function to use hinge-loss instead of a step function. That is, our new objective function to maximise is $$\sum_{r^{\prime} \in L} p_{r^{\prime}}\textrm{min}(1, \frac{ess(r^{\prime})}{minSS} $$
Here we assume that it is acceptable to run our algorithm on samples smaller than $minSS$,  though we still prefer bigger sample sizes upto $minSS$. The other simplification we make is assuming that the sample sizes $n_r$ are real numbers instead of integers. After obtaining our optimal sample sizes, we can round them up to get integer sample sizes. This will increase the memory usage by at most $|U|$, the number of nodes in displayed tree. $|U|$ is usually negligible compared to the memory capacity, or $minSS$.

\subsubsection{Additional optimizations}
There are some additional minor optimizations we can make to reduce the memory cost per sample, allowing us to store more and bigger samples. 
Suppose we have a sample $s$, and say its filter rule $f_s$ has value $v$ in column $c$. Then we know that each tuple $t$ in $T_s$ must also have value $v$ in column $c$, since it is covered by $f_s$. So we do not need to explicitly store the column $c$ value of any tuple in $T_s$. We only need to store the tuple values of columns that have a $\star$ value in $f_s$.
In addition, we may have a tuple occur in multiple samples. Instead of storing the entire tuple repeatedly, we could create a dictionary of common tuples, and only store a pointer to the tuple's dictionary entry in $T_s$.
